Normalization by Evaluation for Call-by-Push-Value and Focused Lambda-Calculus

### 2019-02-08 fscd19/fscd19.tex
[ fscd19 ] started on evaluation function

We are looking for a cover monad $\C$ that
offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{lcl@{\quad}l}
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\A
  & \mbox{case on absurd neutral} \\
\tcase^\C_\Gamma & : & \Ne {(A + B)} \Gamma
  \to \C\,\A\,(\cext \Gamma A)
  \to \C\,\A\,(\cext \Gamma B)
  \to \C\,\A\,\Gamma
  & \mbox{case on neutral} \\
\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
  & \mbox{run the monad ($\tNf$ only)} \\
\end{array}
\]
We shall immediatly construct an instance of such a cover monad,
the free cover monad
$\Cov$ defined inductively by $\treturn[\Cov]$, $\tabort^\Cov$, and
$\tcase^\Cov$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
of sum type $A+B$ and its two branches by the context extensions $A$
and $B$, resp.  Leaves are either ($\tabort$)
labeled by a neutral term of empty type $\tyzero$,
or ($\treturn$) by an element of $\A$.  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind replaces
these leaves by further case trees.  (The uninspiring $\tjoin[\Cov]$
flattens a 2-level case tree into a single one.)
Finally $\trunNf[\Cov]$ is a simple recursion, replacing $\tcase^\Cov$
and $\tabort^\Cov$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

@@

>>>>

We are looking for a cover monad $\C$ that
offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{lcl@{\quad}l}
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\A
  & \mbox{case on absurd neutral} \\
\tcase^\C_\Gamma & : & \Ne {(A + B)} \Gamma
  \to \C\,\A\,(\cext \Gamma A)
  \to \C\,\A\,(\cext \Gamma B)
  \to \C\,\A\,\Gamma
  & \mbox{case on neutral} \\
\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
  & \mbox{run the monad ($\tNf$ only)} \\
\end{array}
\]
To make things concrete,
we shall immediately construct an instance of such a cover monad:
the free cover monad
$\Cov$ defined inductively by $\treturn[\Cov]$, $\tabort^\Cov$, and
$\tcase^\Cov$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
of sum type $A+B$ and its two branches by the context extensions $A$
and $B$, resp.  Leaves are either ($\tabort$)
labeled by a neutral term of empty type $\tyzero$,
or ($\treturn$) by an element of $\A$.  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind replaces
these leaves by further case trees.  (The uninspiring $\tjoin[\Cov]$
flattens a 2-level case tree into a single one.)
Finally $\trunNf[\Cov]$ is a simple recursion, replacing $\tcase^\Cov$
and $\tabort^\Cov$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

@@

We extend the type interpretation pointwise to contexts, \ie,
$\den \cempty = \hatone$ and
$\den{\cext \Gamma A} = \den \Gamma \hattimes \den A$
and obtain a natural projection function from the semantic enironments:
\[
\begin{array}{lcl}
  \tlookup\,(x : A \in \Gamma) & : &  \den \Gamma \todot \den A \\
  \tlookup\;\tzero     & = & \pi_2 \\
  \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
\end{array}
% \begin{array}{lcl}
%   \tlookup & : & A \in \Gamma \to \den \Gamma \todot \den A \\
%   \tlookup\;\tzero     & = & \pi_2 \\
%   \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
% \end{array}
\]
The evaluation function
\fbox{$\dent {t : A \from \Gamma} \_ : \den \Gamma \todot \den A$}
can now be defined by recursion on $t$.
Herein the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
% Herein, $\gamma : \den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
\[
\begin{array}{lcl}
  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
  \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
     \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
        \dent t {(\tren\,\tau\,\gamma,\ a)} \\
  \dent{\tapp\,t\,u} \gamma & = &
     \dent t \gamma\;\tid\;\dent u \gamma \\
  % \dent{\tapp\,(t : A \To B \from \Gamma)\,u} \gamma & = &
  %    \dent t \gamma\;\tid\;\dent u \gamma \\
  \dent{\tunit}\gamma & = & \ttempty \\
  \dent{\tpair\,t_1\,t_2}\gamma & = &
    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
  \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
  \dent{\tcase\,t\,u\,v}\gamma & = & \\
  \dent{\tabort\,t}\gamma & = & \\
\end{array}
\]
